\textbf{ Matching the forecast horizon with the relevant ecological processes }

### 2019-07-22 doc/manuscript.tex
Editoral work on Community Turnover example

In two case studies, we simulate the effects of an increase in temperature on simple systems with known dynamics.
The ``truth'' is represented by a model that is mechanistic for at least one important process, but treat the model as unknown when
analyzing the data and assume that perfectly recovering this model would not be possible in practice. We begin each simulation
under stationary temperature, allowing the system to equilibrate; we call this the baseline phase. We then increase temperature progressively
over a period of time, followed by a second period of stationary, now elevated, temperature. The objective is to forecast the response of the 
system to the temperature increase based on data gathered during the baseline period.

We make forecasts based on two phenomenological models, each representing processes operating at different time scales. 
The first phenomenological model represents what we call a ``spatial approach.'' We correlate the mean environment with the mean of an ecological state or rate 
 across many sites. This is the approach commonly used to predict population distribution or abundance as a function of climate \citep{elith_species_2009} or
 primary production as a function of mean precipitation \citep{Sala1988}. The approach is attractive because it addresses broad spatial scales relevant to 
 management, and because the necessary data are widely available and statistical analyses are straightforward. The approach also has well-known weaknesses.
Criticism often focuses on the assumption that current spatial patterns are in equilibrium with climate (refs?). Another issue is that these models are not dynamic; 
they provide no information about how quickly the system will move from the current state to the predicted, future state. Because the patterns in the data have developed 
over long time spans, they are may not provide good information about dynamic changes on the order of a few years. Moreover, transient dynamics could 
prevent the system from ever reaching the predicted steady state \citep{Urban2012}.

The second model represents what we call a ``temporal approach.'' We correlate interannual variation in an ecological response with interannual variation in the environment, 
typically at just one site. This approach is often used to study population or vital rate fluctuations as a function of weather \citep{dalgleish_climate_2011}, 
or primary production as a function of precipitation \citep{lauenroth_long-term_1992}. The advantages of this approach include is focus on temporal dynamics, meaning it 
can predict rates of change, and validation of the resulting near-term predictions are straightforward. However, models built using this approach typically cover 
a limited spatial extent (but see \citealt{kleinhesselink_response_2018}, others), and ignore slower processes, such as evolutionary adaptation or turnover in community composition, 
that could alter the underlying process at longer time scales and improve forecastability \citep{clark_ecological_2001}.

We compare forecasts from both models to the simulated dynamics to determine how well the two approaches to forecasting perform at different time scales. We also assess the potential for combining the information available in both spatial and temporal patters by combining forecasts from the spatial and temporal approaches using a weighted average and assess how the optimal model weights change over time. Based on the hypothesis that spatial patterns capture longer-term processes we expect the temporal model to best predict short-term dynamics, and the spatial model to best predict long-term dynamics, and that a combination of the two approaches will provide the best forecasts at transitional intermediate time scales.

@@

A common problem for conservation biologists and natural resource managers is anticipating the impact of environmental change on the abundance of 
endangered species, biological invaders, and harvested species. Although the managers may be primarily interested in just the focal species, 
skillful prediction might require considering interactions with many other species, greatly complicating the problem. But at what forecast horizon do
altered species interactions become impossible to ignore? We explored this question using a metacommunity model developed by \cite{alexander_lags_2018}. 
The model describes Lotka-Volterra competitive interactions among plants within sites that are arrayed along an elevation and temperature gradient.
Composition varies along the gradient because of a trade-off between growth rate and cold tolerance: 
cold sites are dominated by slow-growing species that can tolerate low temperatures, while warm sites are dominated by 
fast-growing species that are cold intolerant. Multiple species can coexist within sites because all species experience stronger competition from conspecifics than from heterospecifics. 
Sites are linked by dispersal: a specified fraction of each species' offspring leaves the site where they were produced and reaches all other sites with equal probability. All computer code is available on Github: \texttt{https://github.com/pbadler/space-time-forecast}).

@@

Dynamics during the baseline period revealed very strong spatial patterns across the mean temperature gradient. 
Individual species, including our focal species, showed classic, unimodal ``Whittaker" patterns of abundances across the gradient (Fig. \ref{fig:species-patterns-models}A).
These spatial patterns are the basis for our ``spatial model" of the temperature-biomass relationship for our focal species (Fig. \ref{fig:species-patterns-models}A). 
In contrast to the strong spatial patterns, population and community responses to interannual variation in temperature within sites were weak.
At our focal site in the center of the gradient, the biomass of the focal species was quite insensitive to interannnual variation in temperature (Fig. \ref{fig:species-patterns-models}B). Our ``temporal model" estimates this weak, linear temperature effect, along with strong lag effects of biomass in the previous year. The spatial and temporal statistical
models are described in Appendix \ref{models}.

@@

We forecasted the effect of temperature increase on the focal species' biomass at one location in the center of the temperature gradient 
using the spatial models and the temporal models. 
The predictions from the spatial and temporal model contrasted markedly, with the temporal 
model predicting a large increase in biomass and the spatial model predicting a decrease. Initially, the simulated abundances followed the increase predicted by the 
temporal model, but as faster-growing species colonized and increased in abundance at the study site, the biomass of the focal species then decreased, eventually falling below its baseline level (Fig. \ref{fig:community-forecast-species}).

To combine the spatial and temporal model into a single forecast, we fit a weighting parameter, $\omega$, which varies over time and is bounded between 0 and 1. At any time point, $t$, the combined population forecast is $\omega * T(N_{t-1},K_t) + (1-\omega) * S(K_t) $ where $T$ is the temporal model, which depends on population size, $N$ and expected temperature, $K$, and $S$ is the spatial model, which depends only on $K$ (see Appendix \ref{models} for full description of the approach).
This combined model accurately predicts the simulated dynamics across the full forecast horizon (Fig. \ref{fig:community-forecast-species}) 
because the weights were determined by fitting directly to those data. Unlike our spatial and temporal model forecasts, 
we did not generate out-of-sample predictions from the combined model; it merely provides 
a convenient way to quantify how rapidly dynamics shift from from being dominated by the interannual 
variations captured in the temporal model to being dominated by the steady-state equilibrium captured by the spatial model.
The most rapid shifts in model weights occur during the period when warm-adapted, faster growing species are increasing most rapidly
in abundance (Fig. \ref{fig:community-weights-spp}). A true, combined forecast would require a method to determine the model weights \emph{a priori}.

>>>>

In both of our case studies, we simulate the effects of a hypothetical temperature increase on a simple system with known dynamics.
The ``truth'' is represented by a model that is mechanistic for at least one important process, but we assume that recovering 
this true, process-based model would be impossible in practice (we return to this point in the Discussion). We begin each simulation
under stationary temperature, allowing the system to equilibrate; we call this the baseline phase. We then increase temperature progressively
over a period of time, followed by a second period of stationary, now elevated, temperature. The objective is to forecast the response of the 
system to temperature increase based on data gathered during the baseline period.

We make two forecasts based on two different phenomenological models, each representing processes operating at different time scales. 
The first phenomenological model represents what we call a ``spatial approach.'' We correlate the mean environment with the mean of an ecological state or rate 
 across many sites. This is the approach commonly used to predict population distribution or abundance as a function of climate \citep{elith_species_2009} or
 primary production as a function of mean precipitation \citep{Sala1988}. The approach is attractive because it addresses broad spatial scales relevant to 
 management, and because the necessary data are widely available and statistical analyses are straightforward. The approach also has well-known weaknesses.
Criticism often focuses on the assumption that current spatial patterns are in equilibrium with climate (refs?). Another issue is that these models are not dynamic; 
they provide no information about how quickly the system will move from the current state to the predicted, future state. Because the patterns in the data have developed 
over long time spans, they are unlikely to provide much information about dynamic changes on the order of years to decades. Moreover, transient dynamics could 
prevent the system from ever reaching the predicted steady state \citep{Urban2012}.

The second model represents what we call a ``temporal approach.'' We correlate interannual variation in an ecological response with interannual variation in the environment, 
typically at just one site. This approach is often used to study population or vital rate fluctuations as a function of weather \citep{dalgleish_climate_2011}, 
or primary production as a function of precipitation \citep{lauenroth_long-term_1992}. The advantages of this approach include is focus on temporal dynamics, meaning it 
can predict rates of change, and validation of the resulting near-term predictions are straightforward. However, models built using this approach typically cover 
a limited spatial extent (but see \citealt{kleinhesselink_response_2018}, others), and ignore slower processes, such as evolutionary adaptation or turnover in community composition, 
that could alter the underlying process at longer time scales.

We compare forecasts from both models to the true, simulated dynamics to show that responses to environmental change reflect both kinds of processes, but at different time scales. To quantify this pattern, we combine forecasts from the spatial and temporal approaches with a weighted average and then show how the optimal model weights change over time: we expect the temporal model to best predict short-term dynamics, and the spatial model to best predict long-term dynamics. How rapidly that transition occurs depends on the rate of  ``slow" processes such as species turnover and evolution. 
Code for simulating dynamics and reproducing the analyses in this paper is available on Github \texttt{https://github.com/pbadler/space-time-forecast}) and
archived on Zenodo \texttt{https://zenodo.org}).

@@

A common need for conservation biologists and natural resource managers is anticipating the impact of environmental change on the abundance of 
endangered species, biological invaders, and harvested species. Although the managers may be primarily interested in just the focal species, 
skillful prediction might require considering interactions with many other species, greatly complicating the problem. But at what forecast horizon do
altered species interactions become impossible to ignore? We explored this question using a metacommunity model developed by \cite{alexander_lags_2018}. 
The model describes Lotka-Volterra competitive interactions among plants within sites that are arrayed along an elevation and temperature gradient.
Composition varies along the gradient because of a trade-off between growth rate and cold tolerance: 
cold sites are dominated by slow-growing species that can tolerate low temperatures, while warm sites are dominated by 
fast-growing species that are cold intolerant. Multiple species can coexist within sites because all species experience stronger competition from conspecifics than from heterospecifics. 
Sites are linked by dispersal: a specified fraction of each species' offspring leaves the site where they were produced and reaches all other sites with equal probability.

@@

During the baseline period there were strong spatial patterns across the mean temperature gradient. 
Individual species, including our focal species, showed classic, unimodal ``Whittaker" patterns of abundances across the gradient (Fig. \ref{fig:species-patterns-models}A).
These spatial patterns are the basis for our ``spatial model" of the temperature-biomass relationship for our focal species (Fig. \ref{fig:species-patterns-models}A). 
In contrast to the strong spatial patterns, population and community responses to interannual variation in temperature within sites were weak.
At our focal site in the center of the gradient, the biomass of the focal species was quite insensitive to interannnual variation in temperature (Fig. \ref{fig:species-patterns-models}B). Our ``temporal model" estimates this weak, linear temperature effect, along with strong lag effects of biomass in the previous year. The spatial and temporal statistical
models are described in Appendix \ref{models}.

@@

We forecast the effect of a temperature increase on the focal species' biomass at one location in the center of the temperature gradient 
using both spatial models and the temporal models. 
The predictions from the spatial and temporal models contrasted markedly, with the temporal 
model predicting a large increase in biomass and the spatial model predicting a decrease.Initially, the simulated abundances followed the increase predicted by the 
temporal model, but as faster-growing species colonized and increased in abundance at the focal site, the biomass of the focal species decreased, eventually falling below its baseline level (Fig. \ref{fig:community-forecast-species}).

To combine the spatial and temporal model into a single forecast, we fit a weighting parameter, $\omega$, which varies over time and is bounded between 0 and 1. At any time point, $t$, the combined population forecast is $\omega * T(N_{t-1},K_t) + (1-\omega) * S(K_t) $ where $T$ is the temporal model, which depends on population size, $N$ and expected temperature, $K$, and $S$ is the spatial model, which depends only on $K$ (see Appendix \ref{models} for full description of the approach).
This combined model accurately predicts the simulated dynamics across the full forecast horizon (Fig. \ref{fig:community-forecast-species}), but 
it works so effectively because the weights were determined by fitting directly to those data. Unlike our spatial and temporal model forecasts, 
we did not generate out-of-sample predictions from the combined model; it merely provides 
a convenient way to quantify how rapidly dynamics shift from being dominated by interannual 
variation captured in the temporal model to being dominated by the steady-state equilibrium captured by the spatial model.
A true, combined forecast would require a method to determine the model weights \emph{a priori}.
The most rapid shifts in the model weights occurred during the period when warm-adapted, faster growing species were increasing most rapidly
in abundance (Fig. \ref{fig:community-weights-spp}).
